---
title: "DANA 4820 Project Part 1"
output:
  word_document: default
  html_document: default
date: "2026-07-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(knitr)
library(kableExtra)

# Load the dataset (ensure fHH1.csv is in the same folder as this .Rmd,
# or update the path below)
fHH1 <- read.csv("https://raw.githubusercontent.com/proback/BeyondMLR/master/data/fHH1.csv")
```


**Course: DANA 4820**

**Team members: Nandini Bhatia, Diego Alberto, Khushpreet Kaur, Haruka Sawakami**

# Part (a)
## Summary of the Beyond Multiple Linear Regression, Chapter 4

**Course: DANA 4820**

**Team members: Nandini Bhatia, Diego Alberto, Khushpreet Kaur, Haruka Sawakami**

# Part (a)
# 4. Poisson Regression
## Summary of the Beyond Multiple Linear Regression, Chapter 4

```{r load_packages4, message = FALSE, warning=FALSE}
# Packages required for Chapter 4
#install.packages("mosaic")
#install.packages("pscl")
#install.packages("multcomp")
#install.packages("pander")
library(gridExtra)
library(knitr)
library(kableExtra)
library(mosaic)
library(xtable)
library(pscl) 
library(multcomp)
library(pander)
library(MASS)
library(tidyverse)
library(ggplot2)
```


## 4.1 Learning Objectives
The primary objective of this section is to understand the foundational concepts of Poisson regression and recognize why Simple Linear Regression models are inadequate for modeling count data. Through this chapter, we aim to grasp the model's assumptions for inference, the estimation and interpretation of coefficients using R, and the methods for comparing models through hypothesis testing.


## 4.2 Introduction to Poisson Regression
Poisson regression is specifically designed to model response variables ($Y$) that represent counts, that is the average number of occurrences per unit of time or space. Examples out of four include the annual number of motorcycle deaths in a state or the number of deformed fish in Minnesota lakes over the last decade. Count data has unique characteristics: they are discrete, can take only positive number and theoretically unbounded above (from 0 to $\infty$).

While it might seem to be reasonable to apply a standard Linear Least Squares Regression (LLSR) model, such as $\lambda_i = \beta_0 + \beta_1 x_i$, doing so introduces two critical violations of regression assumptions:

1. **Violation of Boundedness (Negative Values):**<br>
A linear regression line extends infinitely and is guaranteed to predict negative values for certain values of $x_i$. However, the Poisson mean rate parameter ($\lambda_i$) represents an average count and can only take non-negative values.

2. **Violation of Homoscedasticity (Equal Variance):**<br>
In LLSR, the variance of residuals is assumed to be constant across all levels of $X$. For Poisson data, however, as the mean rate ($\lambda$) increases, the variability of the data also increases. This is because a key property of the Poisson distribution is that its mean is equal to its variance ($E(Y) = Var(Y) = \lambda$).


To resolve these issues, instead of modeling $\lambda_i$ directly, we express the natural logarithm of the mean rate, $\log(\lambda_i)$, as a linear function:

$$\log(\lambda_i) = \beta_0 + \beta_1 x_i$$

Taking the exponential of both sides keeps the predicted mean rate $\lambda_i$ positive for any value of $x_i$, which satisfies the mathematical constraint that count data cannot be negative.

Furthermore, unlike the LLSR, the Poisson regression model does not contain a separate error term ($\epsilon$). In the LLSR, $\epsilon$ represents the random variation around the mean. In Poisson regression, a separate error term is unnecessary because the single parameter $\lambda_i$ completely determines both the mean and the variance of the response variable ($Y_i \sim \text{Poisson}(\lambda_i)$).


### 4.2.1 Poisson Regression Assumptions
For a Poisson regression model to use, the following four key assumptions must be satisfied:

1. **Poisson Response:** The response variable must be a count of occurrences that follows a Poisson distribution within a defined unit of time or space.
2. **Independence:** The observations must be independent of one another.
3. **Mean = Variance:** A fundamental property of the Poisson distribution is that its mean is equal to its variance ($E(Y) = Var(Y) = \lambda$). We must check if the means and variances are approximately equal across groups of the explanatory variable.
4. **Linearity:** The log of the mean rate, $\log(\lambda)$, must be a linear function of the explanatory variables ($X$).


### 4.2.2 A Graphical Look at Poisson Regression
```{r, OLSpois, fig.align="center",out.width="60%", fig.cap='Regression models: Linear regression (left) and Poisson regression (right).',echo=FALSE, warning=FALSE, message=FALSE}
## Sample data for graph of OLS normality assumption
## Code from https://stackoverflow.com/questions/31794876/ggplot2-how-to-curve-small-gaussian-densities-on-a-regression-line?rq=1

set.seed(0)
dat <- data.frame(x=(x=runif(10000, 0, 50)),
                  y=rnorm(10000, 10*x, 100))

## breaks: where you want to compute densities
breaks <- seq(0, max(dat$x), len=5)
dat$section <- cut(dat$x, breaks)

## Get the residuals
dat$res <- residuals(lm(y ~ x, data=dat))

## Compute densities for each section, flip the axes, add means 
## of sections.  Note: densities need to be scaled in relation 
## to section size (2000 here)
dens <- do.call(rbind, lapply(split(dat, dat$section), function(x) {
  d <- density(x$res, n=5000)
  res <- data.frame(x=max(x$x)- d$y*1000, y=d$x+mean(x$y))
  res <- res[order(res$y), ]
  ## Get some data for normal lines as well
  xs <- seq(min(x$res), max(x$res), len=5000)
  res <- rbind(res, data.frame(y=xs + mean(x$y),
                x=max(x$x) - 1000*dnorm(xs, 0, sd(x$res))))
  res$type <- rep(c("empirical", "normal"), each=5000)
  res
}))
dens$section <- rep(levels(dat$section), each=10000)

ols_assume <- ggplot(dat, aes(x, y)) +
  geom_point(size = 0.1, alpha = .25) +
  geom_smooth(method="lm", fill=NA, lwd=2) +
  geom_path(data=dens[dens$type=="normal",], 
            aes(x, y, group=section), 
            color="salmon", lwd=1.1) +
  theme_bw() +
  geom_vline(xintercept=breaks, lty=2)

# Now make Poisson regression picture
set.seed(0)
dat <- data.frame(x=(x=runif(1000, 0, 20)),
                  y=rpois(1000, exp(.1*x)))

## breaks: where you want to compute densities
breaks <- seq(2, max(dat$x), len=5)
dat$section <- cut(dat$x, breaks)

## Get the residuals
dat$res <- dat$y - .1*dat$x

## Compute densities for each section, flip the axes, add means
## of sections.  Note: densities need to be scaled in relation 
## to section size
dens <- do.call(rbind, lapply(split(dat, dat$section), function(x) {
  d <- density(x$res, n=500)
  res <- data.frame(x=max(x$x)- d$y*10, y=d$x+mean(x$y))
  res <- res[order(res$y), ]
  ## Get some data for poisson lines as well
  xs <- seq(min(x$y), max(x$y), len=500)
  res <- rbind(res, data.frame(y=xs,
          x=max(x$x) - 10*dpois(round(xs), exp(.1*max(x$x)))))
  res$type <- rep(c("empirical", "poisson"), each=500)
  res
}))
dens$section <- rep(levels(dat$section), each=1000)

pois_assume <- ggplot(dat, aes(x, jitter(y, .25))) +
  geom_point(size = 0.1) +
  geom_smooth(method="loess", fill=NA, lwd=2) +
  geom_path(data=dens[dens$type=="poisson",], 
            aes(x, y, group=section), 
            color="salmon", lwd=1.1) +
  theme_bw() + ylab("y") + xlab("x") +
  geom_vline(xintercept=breaks, lty=2)

grid.arrange(ols_assume, pois_assume, ncol = 2)
```

Figure illustrates a comparison of the LLSR model for inference to Poisson regression using a log function of $\lambda$.

Before fitting a Poisson regression model, we should visually evaluate our data to see if the assumptions hold. In standard linear regression, we look for a straight line in a scatterplot . However, for a Poisson regression model:

1. **We do not expect a linear relationship between $X$ and the raw counts $Y$.** Instead, if the relationship follows a Poisson distribution, a scatterplot of the explanatory variable ($X$) against the response variable ($Y$) should show an exponential curve rather than a straight line.
2. **We look for a linear relationship between $X$ and the log of the mean response.** To visually check the linearity assumption ($\log(\lambda) = \beta_0 + \beta_1 X$), we can break a continuous explanatory variable into discrete groups and calculate the mean for each group. 
3. **The plot of the log means against $X$ should look approximately linear.** If the linearity assumption is satisfied, plotting the natural logarithm of these means against the average $X$ values for each group should show a reasonably straight line.


## 4.3 Case Study Overview
In this section, we overview the three case studies introduced in the textbook, each based on real data and real questions to illustrate different aspects of Poisson regression modeling:

1. **Household Size in the Philippines:** Introduces the basic idea of regression with a Poisson response along with its structural assumptions, adding a quadratic term to determine an optimal size, and methods of model comparison.
2. **Campus Crime:** Introduces offsets to account for sampling effort and overdispersion when actual variability exceeds what is expected.
3. **Weekend Drinking:** Uses a modification of a Poisson model to account for more zeros than expected.

*Note: Since the topics covered with campus crime and weekend drinking case studies are beyond the scope of our course, we will focus exclusively on the first case study: Household Size in the Philippines.*


## 4.4 Case Study: Household Size in the Philippines

How many people live in your home? The number of people sharing a house 
differs from country to country and region to region. International 
agencies use household size when determining the needs of populations, 
so understanding what drives these differences is practically important.

In this case study we look at household size in the Philippines and try 
to answer two main questions:

1. *At what age are heads of households in the Philippines most likely 
to find the largest number of people in their household?*
2. *Is this association between household size and the head's age similar 
for poorer households?*

To answer these, we look at the relationship between household size and 
the age of the household head, while also controlling for where the 
household is located and what type of roof material they have (used as 
a proxy for income level).

The data comes from the 2015 Family Income and Expenditure Survey (FIES), 
conducted by the Philippine Statistics Authority (PSA). Our dataset is a 
subset of 1,500 observations drawn from 40,000 total, focusing on five 
regions: Central Luzon, Metro Manila, Ilocos Region, Davao Region, and 
Visayas.

Our response variable $Y$ is the number of household members other than 
the head of the household. Since this is a count variable — it can only 
take values 0, 1, 2, 3... — we use Poisson regression where the 
parameter of interest is $\lambda$, the average number of non-head 
members per household.

### Chapter 4.4.1
### Chapter 4.4.2
### Chapter 4.4.3

## 4.4.4 Using Deviances to Compare Models

Our objective here is to assess if the age is a significant predictor for the household size, a drop in deviance test was conducted to compare the null model to the model incorporating age as linear predictor.  
The hypotheses are formally stated as follows:

$$H_0: \log(\lambda) = \beta_0 \quad (\beta_1 = 0, \text{ age is not a significant predictor})$$

$$H_A: \log(\lambda) = \beta_0 + \beta_1\text{age} \quad (\beta_1 \neq 0, \text{ age is a significant predictor})$$
```{r}
# Null model
model0 <- glm(total ~ 1, family = poisson, data = fHH1)

# Model with age (linear)
modela <- glm(total ~ age, family = poisson, data = fHH1)

# Drop-in-deviance test
drop_in_dev_1 <- anova(model0, modela, test = "Chisq")
drop_in_dev_1
```

The null model, which contains no predictors, has a residual deviance of 2362 on 1499 degrees of freedom. When included the age as a predictor, the residual deviance decreased to 2337 on 1498 degrees of freedom, this results in a drop in deviance of 25.4 with one degree of freedom. Then we evaluate this against a chi-quare distribution with one degree of freedom, resulting in a very tiny p-value ($4.66 \times
10^{-7}$) that is below the significance threshold of 0.05
Interpretation: We reject $H_0$. There is statistically significant evidence that age
is a meaningful predictor of household size, and therefore the model with
age is preferred over the null model 

## 4.4.6 Second Order Model

We already showed that age matters for household size. Then when we looked at our exploratory graphs earlier(section 4.4.2), the relationship wasn't a straight line - it when up, peaked and then came down. To determine whether the quadratic term significantly improves model fit, a drop-in-deviance test is conducted comparing the linear model to the quadratic model. The hypotheses are
formally stated as follows:

$$H_0: \log(\lambda) = \beta_0 + \beta_1\text{age} \quad (\beta_2 = 0, \text{ quadratic term is not needed})$$

$$H_A: \log(\lambda) = \beta_0 + \beta_1\text{age} + \beta_2\text{age}^2 \quad (\beta_2 \neq 0, \text{ quadratic term is needed})$$

```{r}
# Add quadratic term
fHH1 <- fHH1 %>% mutate(age2 = age * age)

# Fit quadratic model
modela2 <- glm(total ~ age + age2, family = poisson, data = fHH1)
summary(modela2)

# Drop-in-deviance test: linear vs quadratic
drop_in_dev_2 <- anova(modela, modela2, test = "Chisq")
drop_in_dev_2

# Calculate optimal age
beta1 <- coef(modela2)["age"]
beta2 <- coef(modela2)["age2"]
optimal_age <- -beta1 / (2 * beta2)
cat("Optimal age:", round(optimal_age, 2))

# Maximum estimated household size
max_log_total <- coef(modela2)[1] + beta1 * optimal_age +
                 beta2 * optimal_age^2
cat("Max mean household size:", round(exp(max_log_total), 3))
```



$$\log(\text{mean numHouse}) = -0.333 + 0.071 \cdot \text{age} - 0.00071 \cdot \text{age}^2$$

The positive coefficient associated with the linear age  ($\hat{\beta}_1
= 0.071$, $p < 0.001$) indicates that household size initially increases
with the age of the household head. The negative coefficient associated
with the quadratic term ($\hat{\beta}_2 = -0.00071$, $p < 0.001$) confirms
the presence of a concave relationship, in other words household size reaches a
maximum at a certain age before subsequently declining. We can expect that household size grows as children are born and raised, then decreases as children leave the household.

We then asked: does adding age^2 actually make the model better or not? We compared the simpler model to the more complex model(age+age^2). The deviance dropped by 136.1 and the p-value is almost zero. Therefore  we reject $H_0$and conclude that the inclusion of the quadratic term significantly improves model fit, and the quadratic model is strongly preferred over the linear model.

To determine the age at which household size is estimated to be at its
maximum, the derivative of the fitted model with respect to age was set
equal to zero:

$$\frac{d}{d(\text{age})}\log(\text{total}) = 0.071 - 2(0.00071)\cdot\text{age} = 0$$
$$\text{age} = 0.071 / 0.00142 \approx 50.04 \text{ years}$$

To find exactly where the peak is, we use a basic calculus trick: at the very top of a curve, the slope is flat (zero). So we take the derivative of our equation, set it to zero, and solve for age. That gives us age ≈ 50.04.

Then we plug that age back into the original equation to find out how big the household is at that peak. That gives us a log value of 1.441, and converting that back to a normal number (using e^1.441) gives us about 4.23 people.

Interpretation: Household size peaks when the head of household is around 50 years old, with about 4.2 extra people living in the home at that point — then it starts declining again.

## 4.4.7 Adding a Covariate

So far our model only uses age to predict household size. But it's reasonable to think that where someone lives might also affect how many people are in their household. So we wanted to test: after we've already accounted for age, does location also matter?

Location isn't a number, it's a category (Central Luzon, Davao, Ilocos, Metro Manila, Visayas), so R can't just drop it into the equation as-is. Instead, R picks one category as the "baseline" to compare everything else to — here it picked Central Luzon — and creates a 0/1 indicator for each of the other four regions. So basically, the model is asking "compared to Central Luzon, is this region higher or lower?" for each region separately. The hypotheses are formally stated 
as follows:

$$H_0: \log(\lambda) = \beta_0 + \beta_1\text{age} + \beta_2\text{age}^2 \quad (\text{location does not improve the model})$$

$$H_A: \log(\lambda) = \beta_0 + \beta_1\text{age} + \beta_2\text{age}^2 + \beta_3\text{location} \quad (\text{location improves the model})$$



```{r}
# Fit quadratic model with location
modela2L <- glm(total ~ age + age2 + location,
                family = poisson, data = fHH1)
summary(modela2L)

# Drop-in-deviance test: quadratic vs quadratic + location
drop_in_dev_3 <- anova(modela2, modela2L, test = "Chisq")
drop_in_dev_3

# Rate ratios and confidence intervals
exp(coef(modela2L))
exp(confint(modela2L))
```


The fitted model incorporating location is expressed as follows:

$$\log(\text{total}) = -0.384 + 0.070 \cdot \text{age} - 0.00070 \cdot \text{age}^2 + 0.061 \cdot \text{IlocosRegion} + 0.054 \cdot \text{MetroManila} + 0.112 \cdot \text{Visayas} - 0.019 \cdot \text{DavaoRegion}$$

The coefficients for age and age$^2$ remained stable in magnitude and
statistical significance following the inclusion of location, suggesting
that the quadratic relationship between age and household size is robust
to the addition of this covariate.

Reading the region coefficients individually:

- **Visayas** ($\hat{\beta} = 0.112$, $p = 0.007$): Households here are 
  about 11.9% larger than Central Luzon, on average, after controlling 
  for age, and this difference is statistically significant.
- **Ilocos Region** ($\hat{\beta} = 0.061$, $p = 0.25$): Slightly larger 
  than Central Luzon, but not statistically significant — this could 
  simply be due to random variation.
- **Metro Manila** ($\hat{\beta} = 0.054$, $p = 0.25$): Similarly slightly 
  larger than Central Luzon, but also not statistically significant.
- **Davao Region** ($\hat{\beta} = -0.019$, $p = 0.72$): Slightly smaller 
  than Central Luzon, but again not statistically significant.

So, when considered individually, only Visayas stands out as truly 
different from Central Luzon.

When a categorical variable like location creates multiple dummy variables, you can't just look at each one separately and decide if "location matters." You have to test whether location as a whole — all 4 regions together — improves the model. That's exactly what the drop-in-deviance test does.

We compared the model without location to the model with location, and the deviance dropped by 13.14 with a p-value of 0.011 — below 0.05. So even though 3 out of 4 regions weren't individually significant, as a group, location does meaningfully improve the model. Therefore, we 
reject $H_0$ and conclude that location, as a whole, is a statistically 
significant predictor of household size after controlling for age, and 
it is retained in the final model.




