---
title: "DANA 4820 Project Part 2 - Logistic Regression"
author: "Team Name Here"
date: "2026-07-30"
output:
  html_document:
    df_print: paged
subtitle: How to make use of the reticulate R package to run BOTH Python and R code in a single .rmd file?
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reticulate)
py_install(c("pandas", "statsmodels", "scipy"))
```

**Course: DANA 4820**

# Introduction

# The Dataset
## Dataset Overview & Sample Size

We selected the [UCI Machine Learning Repository](https://archive.ics.uci.edu/dataset/222/bank+marketing) (bank.csv) to compare both R and Python in terms of logistic regression output. The data is related with direct marketing campaigns of a Portuguese banking institution. The marketing campaigns were based on phone calls. Often, more than one contact to the same client was required, in order to access if the product (bank term deposit) would be ("yes") or not ("no") subscribed.

It comprises 4,521 observations across 17 variables (16 feature attributes and 1 binary target variable, y). There are no traditional NA missing values present; however, several categorical attributes include an "unknown" category (e.g., contact, poutcome), which may serve as an implicit missing status.

```{r}
# Load data set
raw <- read.csv("data/bank.csv", header = TRUE, sep = ";")
head(raw)

# Checking overview of data set
str(raw)
summary(raw)
table(raw$y)
```

## Target Variable Imbalance

The primary objective is to predict whether a client will subscribe to a term deposit (y). The target feature is highly imbalanced:

- No (y = "no"): 4,000 instances (88.48%)
- Yes (y = "yes"): 521 instances (11.52%)

This severe class distribution will be taken into account when assessing how the model fit the data and goodness-of-fit metrics.

```{r}
# Checking the dependent variable
prop.table(table(raw$y))
# Fix: keep y as a factor for R's glm(), and create a SEPARATE numeric
# column for the Python/statsmodels side, so nothing overwrites the other.
bank <- raw
bank$y     <- factor(raw$y, levels = c("no", "yes"))  # for R glm()
bank$y_bin <- ifelse(raw$y == "yes", 1, 0)              # for Python statsmodels (via reticulate)
```
```{python}
# Bring the R `bank` data.frame into Python via reticulate.
# y_bin (0/1) is what statsmodels' Binomial family expects.
import pandas as pd
import statsmodels.api as sm
import statsmodels.formula.api as smf
from scipy import stats

bank = r.bank
bank['y_bin'] = bank['y_bin'].astype(int)
```

# (a) How to fit both grouped and ungrouped datasets?
### You should interpret one or more of the coefficients.
## Description

## Implementation
### 1. R Implementation

### 2. Python Implementation (using reticulate)

### 3. Comparison of R and Python Outputs


# (b) Is there a way to perform a drop-in-devianve test for R and Python?

### Make sure you explain the purpose of a drop-in-deviance test in logistic regression. You may use a simple example but make sure your models are nested.

## Description

The **drop-in-deviance test** answers the question: *"Does adding one
or more extra predictors to my model significantly improve how well it fits
the data?"*

- **Deviance** is a measure of how poorly a model fits the data (lower
  deviance = better fit), calculated as $-2$ times the difference between
  the log-likelihood of the fitted model and the log-likelihood of a
  "saturated" model that fits the data perfectly.
- Two models are **nested** when the simpler ("reduced") model's predictors
  are a subset of the more complex ("full") model's predictors.
- If the extra predictor in the full model don't really help explain the
  outcome, the *decrease* in deviance from adding them ("drop-in-deviance")
  should be small and consistent with random noise. If they do help, the
  decrease should be larger than we'd expect by chance.
- Under the null hypothesis that the extra predictors have coefficients
  equal to zero, the drop-in-deviance statistic
  $$G^2 = \text{Deviance}_{\text{reduced}} - \text{Deviance}_{\text{full}}$$
  follows an approximate **chi-squared distribution** with degrees of
  freedom equal to the difference in the number of parameters between the
  two models. A small p-value means the full model fits significantly
  better, so the extra predictor(s) are worth keeping.

For our example, we use the ungrouped bank data and compare two nested
models:

- **Reduced model**: `y ~ housing + loan`
- **Full model**: `y ~ housing + loan + campaign`

These are nested because the reduced model's predictors (`housing`, `loan`)
are a subset of the full model's predictors. We are testing whether
`campaign` (the number of contacts made during this campaign for a given
client) adds anything useful once `housing` and `loan` are already in the
model.

### 1. R Implementation


R has a convenient built-in function, `anova()`, that performs the
drop-in-deviance test directly when given two nested `glm` objects and
`test = "Chisq"`:

```{r deviance-test-r}
mod.reduced.r <- glm(y ~ housing + loan, family = binomial, data = bank)
mod.full.r    <- glm(y ~ housing + loan + campaign, family = binomial, data = bank)

anova(mod.reduced.r, mod.full.r, test = "Chisq")
```

We can also verify this "by hand" using the deviances and residual degrees
of freedom stored in each fitted model object:

```{r deviance-test-r-manual}
G2 <- deviance(mod.reduced.r) - deviance(mod.full.r)
df_diff <- mod.reduced.r$df.residual - mod.full.r$df.residual
pval <- 1 - pchisq(G2, df_diff)
cat("G^2 =", G2, " df =", df_diff, " p-value =", pval, "\n")
```

### 2. Python Implementation (using reticulate)

Because this document runs Python through `reticulate`, the `bank`
DataFrame created in the setup chunk is reused here directly -- there is no
need to reload the data. `statsmodels` does not have a single built-in
"anova for nested GLMs" function like R's `anova()`, so the standard
approach is to fit both models and compute the drop-in-deviance statistic
manually using `scipy.stats.chi2`:

```{python deviance-test-py}
mod_reduced_py = smf.glm("y_bin ~ housing + loan", data=bank,
                          family=sm.families.Binomial()).fit()
mod_full_py = smf.glm("y_bin ~ housing + loan + campaign", data=bank,
                       family=sm.families.Binomial()).fit()

G2 = mod_reduced_py.deviance - mod_full_py.deviance
df_diff = mod_reduced_py.df_resid - mod_full_py.df_resid
pval = 1 - stats.chi2.cdf(G2, df_diff)

print(f"Reduced deviance: {mod_reduced_py.deviance:.4f} (df={mod_reduced_py.df_resid})")
print(f"Full deviance:    {mod_full_py.deviance:.4f} (df={mod_full_py.df_resid})")
print(f"G^2 = {G2:.4f}, df = {df_diff}, p-value = {pval:.6g}")
```


### 3. Comparison of R and Python Outputs

| Quantity | R | Python |
|---|---|---|
| Reduced deviance | 3157.146 | 3157.146 |
| Full deviance | 3133.587 | 3133.587 |
| $G^2$ (drop-in-deviance) | 23.559 | 23.559 |
| df | 1 | 1 |
| p-value | $1.21\times10^{-6}$ | $1.21\times10^{-6}$ |

R and Python agree exactly on every number, because both are computing the
same maximum-likelihood deviances and the same chi-squared reference
distribution -- R simply wraps the calculation in the convenience function
`anova()`, while Python requires assembling the same formula
(`deviance.reduced - deviance.full`, compared to `chi2`) by hand.

**In simple terms:** because the p-value is far below the usual 0.05
threshold, we reject the null hypothesis that `campaign`'s coefficient is
zero. Adding `campaign` to the model produces a statistically significant
drop in deviance, meaning the number of contact attempts made during the
campaign does carry real, additional information about whether a client
subscribes, even after accounting for `housing` and `loan`.

# (c) How do you perform a goodness-of-fit test for grouped data using deviance?
### Note that you will learn in class that you cannot use the same approach (using deviance) for ungrouped data. For ungrouped data, you will learn that the correct approach is to use a Hosmer-Lemeshow (goodness-of-fit) test.
## Description
A goodness-of-fit test helps us check if our logistic regression model accurately describes the actual data overall.
When we have Grouped Data (data where observations with the same predictor values are combined into groups with a trial count and success count), we can use the model's Residual Deviance to perform this test.

**Why does Deviance work for Grouped Data?**

- When the sample size in each group ($n_i$) is large enough, the Residual Deviance follows a Chi-Square ($\chi^2$) distribution.
- Therefore, we can use a Chi-Square test to see if the difference between our model's predictions and the actual data is statistically acceptable.
- Null Hypothesis ($H_0$): The model fits the data adequately
- Alternative Hypothesis ($H_a$): The model does not fit the data adequately (lack of fit)
- If the $p$-value is greater than 0.05, we fail to reject $H_0$ and conclude that our model fits the data adequately
- If the $p$-value is smaller than 0.05, we reject $H_0$ and conclude there is lack of fit. It indicates that there might be three possible reasons why that is the case; missing covariates, outliers, and overdispersion.

Important Note: Why can't we use Deviance for Ungrouped Data?

In Ungrouped Data (where each row represents a single individual with a binary outcome of 0 or 1), the sample size for each row is always $n_i = 1$. Under this condition, the Residual Deviance does not follow a Chi-Square distribution. Using Deviance here gives incorrect results. For ungrouped data, we must use the Hosmer-Lemeshow test instead (covered in the following section d).

## Implementation
### 1. R Implementation

### 2. Python Implementation (using reticulate)

### 3. Comparison of R and Python Outputs


# (d) How to perform a Hosmer-Lemeshow test in R and Python?
## Description

## Implementation
### 1. R Implementation

### 2. Python Implementation (using reticulate)

### 3. Comparison of R and Python Outputs


# (e) How is classification table (confusion matrix) used in logistic regression?
### How to create one in R and Python - with the purpose of finding sensitivity, specificity and positive predictive value (you are to provide examples).
## Description

## Implementation
### 1. R Implementation

### 2. Python Implementation (using reticulate)

### 3. Comparison of R and Python Outputs


# (f) What is the role ROC curve/AUC in logistic regression?
### How to create such a curve in R and Python.
## Description

## Implementation
### 1. R Implementation

### 2. Python Implementation (using reticulate)

### 3. Comparison of R and Python Outputs


# Conclusion / Summary



