---
title: "DANA 4820 Project Part 2 - Logistic Regression"
author: "Team Name Here"
date: "2026-07-30"
output:
  html_document:
    df_print: paged
subtitle: How to make use of the reticulate R package to run BOTH Python and R code in a single .rmd file?
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Course: DANA 4820**

# Introduction

# The Dataset
We selected the [UCI Machine Learning Repository](https://archive.ics.uci.edu/dataset/222/bank+marketing) (bank.csv) to compare both R and Python in terms of logistic regression output.

```{r}
raw <- read.csv("data/bank.csv", header = TRUE, sep = ";")
head(raw)

# Checking overall structure of data set
str(raw)
summary(raw)

# The composition ratio of the dependent variable (y)
table(raw$y)
prop.table(table(raw$y))

bank <- raw
```


# (a) How to fit both grouped and ungrouped datasets?
### You should interpret one or more of the coefficients.
## Description

## Implementation
### 1. R Implementation

### 2. Python Implementation (using reticulate)

### 3. Comparison of R and Python Outputs


# (b) Is there a way to perform a drop-in-devianve test for R and Python?
### Make sure you explain the purpose of a drop-in-deviance test in logistic regression. You may use a simple example but make sure your models are nested.
## Description

## Implementation
### 1. R Implementation

### 2. Python Implementation (using reticulate)

### 3. Comparison of R and Python Outputs


# (c) How do you perform a goodness-of-fit test for grouped data using deviance?
### Note that you will learn in class that you cannot use the same approach (using deviance) for ungrouped data. For ungrouped data, you will learn that the correct approach is to use a Hosmer-Lemeshow (goodness-of-fit) test.
## Description
A goodness-of-fit test helps us check if our logistic regression model accurately describes the actual data overall.
When we have Grouped Data (data where observations with the same predictor values are combined into groups with a trial count and success count), we can use the model's Residual Deviance to perform this test.

**Why does Deviance work for Grouped Data?**

- When the sample size in each group ($n_i$) is large enough, the Residual Deviance follows a Chi-Square ($\chi^2$) distribution.
- Therefore, we can use a Chi-Square test to see if the difference between our model's predictions and the actual data is statistically acceptable.
- Null Hypothesis ($H_0$): The model fits the data adequately
- Alternative Hypothesis ($H_a$): The model does not fit the data adequately (lack of fit)
- If the $p$-value is greater than 0.05, we fail to reject $H_0$ and conclude that our model fits the data adequately
- If the $p$-value is smaller than 0.05, we reject $H_0$ and conclude there is lack of fit. It indicates that there might be three possible reasons why that is the case; missing covariates, outliers, and overdispersion.

Important Note: Why can't we use Deviance for Ungrouped Data?

In Ungrouped Data (where each row represents a single individual with a binary outcome of 0 or 1), the sample size for each row is always $n_i = 1$. Under this condition, the Residual Deviance does not follow a Chi-Square distribution. Using Deviance here gives incorrect results. For ungrouped data, we must use the Hosmer-Lemeshow test instead (covered in the following section d).

## Implementation
### 1. R Implementation

### 2. Python Implementation (using reticulate)

### 3. Comparison of R and Python Outputs


# (d) How to perform a Hosmer-Lemeshow test in R and Python?
## Description

## Implementation
### 1. R Implementation

### 2. Python Implementation (using reticulate)

### 3. Comparison of R and Python Outputs


# (e) How is classification table (confusion matrix) used in logistic regression?
### How to create one in R and Python - with the purpose of finding sensitivity, specificity and positive predictive value (you are to provide examples).
## Description

## Implementation
### 1. R Implementation

### 2. Python Implementation (using reticulate)

### 3. Comparison of R and Python Outputs


# (f) What is the role ROC curve/AUC in logistic regression?
### How to create such a curve in R and Python.
## Description

## Implementation
### 1. R Implementation

### 2. Python Implementation (using reticulate)

### 3. Comparison of R and Python Outputs


# Conclusion / Summary



