---
title: "DANA 4820 Project Part 2 - Logistic Regression"
author: "Team Name Here"
date: "2026-07-30"
output:
  html_document:
    df_print: paged
subtitle: How to make use of the reticulate R package to run BOTH Python and R code in a single .rmd file?
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reticulate)
py_install(c("pandas", "statsmodels", "scipy"))
```

**Course: DANA 4820**

# Introduction

# The Dataset
## Dataset Overview & Sample Size

We selected the [UCI Machine Learning Repository](https://archive.ics.uci.edu/dataset/222/bank+marketing) (bank.csv) to compare both R and Python in terms of logistic regression output. The data is related with direct marketing campaigns of a Portuguese banking institution. The marketing campaigns were based on phone calls. Often, more than one contact to the same client was required, in order to access if the product (bank term deposit) would be ("yes") or not ("no") subscribed.

It comprises 4,521 observations across 17 variables (16 feature attributes and 1 binary target variable, y). There are no traditional NA missing values present; however, several categorical attributes include an "unknown" category (e.g., contact, poutcome), which may serve as an implicit missing status.

```{r}
# Load data set
raw <- read.csv("data/bank.csv", header = TRUE, sep = ";")
head(raw)

# Checking overview of data set
str(raw)
summary(raw)
table(raw$y)
```

## Target Variable Imbalance

The primary objective is to predict whether a client will subscribe to a term deposit (y). The target feature is highly imbalanced:

- No (y = "no"): 4,000 instances (88.48%)
- Yes (y = "yes"): 521 instances (11.52%)

This severe class distribution will be taken into account when assessing how the model fit the data and goodness-of-fit metrics.

```{r}
# Checking the dependent variable
prop.table(table(raw$y))
# Fix: keep y as a factor for R's glm(), and create a SEPARATE numeric
# column for the Python/statsmodels side, so nothing overwrites the other.
bank <- raw
bank$y     <- factor(raw$y, levels = c("no", "yes"))  # for R glm()
bank$y_bin <- ifelse(raw$y == "yes", 1, 0)              # for Python statsmodels (via reticulate)
```
```{python}
# Bring the R `bank` data.frame into Python via reticulate.
# y_bin (0/1) is what statsmodels' Binomial family expects.
import pandas as pd
import statsmodels.api as sm
import statsmodels.formula.api as smf
from scipy import stats

bank = r.bank
bank['y_bin'] = bank['y_bin'].astype(int)
```

# (a) How to fit both grouped and ungrouped datasets?
### You should interpret one or more of the coefficients.
## Description
Logistic regression data can be represented in two equivalent forms:

- **Ungrouped (Bernoulli) form**: every row is a single subject with a
  0/1 outcome. This is how `bank` is stored -- each row is one client,
  and `y_bin` is 0 or 1 for that client only.
- **Grouped (Binomial) form**: rows with identical covariate patterns
  are collapsed into a single row storing the number of trials $n_i$
  (how many subjects share that covariate pattern) and the number of
  successes $s_i$ (how many of them had $y = 1$).

Both forms use the **same likelihood** and therefore produce **identical
coefficient estimates, standard errors, and $z$-statistics** -- collapsing
identical rows together does not change the information in the data, it
just changes how that information is bookkept. What *does* change is the
residual degrees of freedom and the deviance, since the grouped model has
one residual per covariate pattern instead of one residual per subject
(this is exactly why grouped deviance can be used for a goodness-of-fit
test in part (c), while ungrouped deviance cannot).

For this example we fit `y ~ housing + loan`. Both `housing` and `loan`
are binary ("yes"/"no"), so there are only $2 \times 2 = 4$ unique
covariate patterns -- a natural, small example to collapse into grouped
form.

## Implementation
### 1. R Implementation
**Ungrouped fit** -- one row per client:

```{r ungrouped-r}
mod.ungrouped.r <- glm(y ~ housing + loan, family = binomial, data = bank)
summary(mod.ungrouped.r)
```

**Grouped fit** -- collapse the 4,521 rows down to the 4 unique
`housing` x `loan` combinations, storing trials ($n$) and successes:

```{r grouped-r}
grouped.r <- aggregate(y_bin ~ housing + loan, data = bank,
                        FUN = function(x) c(success = sum(x), n = length(x)))
grouped.r <- do.call(data.frame, grouped.r)
names(grouped.r) <- c("housing", "loan", "success", "n")
grouped.r$failure <- grouped.r$n - grouped.r$success
grouped.r

mod.grouped.r <- glm(cbind(success, failure) ~ housing + loan,
                      family = binomial, data = grouped.r)
summary(mod.grouped.r)
```

Comparing coefficients directly confirms they match:

```{r compare-coef-r}
round(cbind(ungrouped = coef(mod.ungrouped.r),
            grouped   = coef(mod.grouped.r)), 6)
```
### 2. Python Implementation (using reticulate)
**Ungrouped fit**:

```{python ungrouped-py}
mod_ungrouped_py = smf.glm("y_bin ~ housing + loan", data=bank,
                            family=sm.families.Binomial()).fit()
print(mod_ungrouped_py.summary())
```

**Grouped fit** -- aggregate with pandas, then fit on the
success/failure counts (statsmodels accepts a two-column response the
same way R's `cbind()` does):

```{python grouped-py}
grouped_py = (bank.groupby(["housing", "loan"])["y_bin"]
                    .agg(success="sum", n="count")
                    .reset_index())
grouped_py["failure"] = grouped_py["n"] - grouped_py["success"]
grouped_py

mod_grouped_py = smf.glm("success + failure ~ housing + loan",
                          data=grouped_py,
                          family=sm.families.Binomial()).fit()
print(mod_grouped_py.summary())
```
### 3. Comparison of R and Python Outputs

| Coefficient | Ungrouped (R) | Grouped (R) | Ungrouped (Python) | Grouped (Python) |
|---|---|---|---|---|
| Intercept | matches across all four | matches across all four | matches across all four | matches across all four |
| `housing[yes]` | matches across all four | matches across all four | matches across all four | matches across all four |
| `loan[yes]` | matches across all four | matches across all four | matches across all four | matches across all four |

All four fits (R ungrouped, R grouped, Python ungrouped, Python grouped)
return numerically identical coefficients, because grouping rows with
identical predictor values does not change the likelihood being
maximized -- only the bookkeeping of trials versus individual 0/1 rows
changes.

**Interpretation of a coefficient:** the coefficient on `housing[yes]`
is negative and highly significant. Exponentiating it gives an odds
ratio below 1, so holding `loan` fixed, a client who has a **housing
loan** has lower odds of subscribing to the term deposit than a client
without a housing loan. Similarly, the coefficient on `loan[yes]` is
negative, meaning clients who already carry a personal loan have lower
odds of subscribing than clients without one, holding `housing` fixed.

# (b) Is there a way to perform a drop-in-devianve test for R and Python?

### Make sure you explain the purpose of a drop-in-deviance test in logistic regression. You may use a simple example but make sure your models are nested.

## Description

The **drop-in-deviance test** answers the question: *"Does adding one
or more extra predictors to my model significantly improve how well it fits
the data?"*

- **Deviance** is a measure of how poorly a model fits the data (lower
  deviance = better fit), calculated as $-2$ times the difference between
  the log-likelihood of the fitted model and the log-likelihood of a
  "saturated" model that fits the data perfectly.
- Two models are **nested** when the simpler ("reduced") model's predictors
  are a subset of the more complex ("full") model's predictors.
- If the extra predictor in the full model don't really help explain the
  outcome, the *decrease* in deviance from adding them ("drop-in-deviance")
  should be small and consistent with random noise. If they do help, the
  decrease should be larger than we'd expect by chance.
- Under the null hypothesis that the extra predictors have coefficients
  equal to zero, the drop-in-deviance statistic
  $$G^2 = \text{Deviance}_{\text{reduced}} - \text{Deviance}_{\text{full}}$$
  follows an approximate **chi-squared distribution** with degrees of
  freedom equal to the difference in the number of parameters between the
  two models. A small p-value means the full model fits significantly
  better, so the extra predictor(s) are worth keeping.

For our example, we use the ungrouped bank data and compare two nested
models:

- **Reduced model**: `y ~ housing + loan`
- **Full model**: `y ~ housing + loan + campaign`

These are nested because the reduced model's predictors (`housing`, `loan`)
are a subset of the full model's predictors. We are testing whether
`campaign` (the number of contacts made during this campaign for a given
client) adds anything useful once `housing` and `loan` are already in the
model.

### 1. R Implementation


R has a convenient built-in function, `anova()`, that performs the
drop-in-deviance test directly when given two nested `glm` objects and
`test = "Chisq"`:

```{r deviance-test-r}
mod.reduced.r <- glm(y ~ housing + loan, family = binomial, data = bank)
mod.full.r    <- glm(y ~ housing + loan + campaign, family = binomial, data = bank)

anova(mod.reduced.r, mod.full.r, test = "Chisq")
```

We can also verify this "by hand" using the deviances and residual degrees
of freedom stored in each fitted model object:

```{r deviance-test-r-manual}
G2 <- deviance(mod.reduced.r) - deviance(mod.full.r)
df_diff <- mod.reduced.r$df.residual - mod.full.r$df.residual
pval <- 1 - pchisq(G2, df_diff)
cat("G^2 =", G2, " df =", df_diff, " p-value =", pval, "\n")
```

### 2. Python Implementation (using reticulate)

Because this document runs Python through `reticulate`, the `bank`
DataFrame created in the setup chunk is reused here directly -- there is no
need to reload the data. `statsmodels` does not have a single built-in
"anova for nested GLMs" function like R's `anova()`, so the standard
approach is to fit both models and compute the drop-in-deviance statistic
manually using `scipy.stats.chi2`:

```{python deviance-test-py}
mod_reduced_py = smf.glm("y_bin ~ housing + loan", data=bank,
                          family=sm.families.Binomial()).fit()
mod_full_py = smf.glm("y_bin ~ housing + loan + campaign", data=bank,
                       family=sm.families.Binomial()).fit()

G2 = mod_reduced_py.deviance - mod_full_py.deviance
df_diff = mod_reduced_py.df_resid - mod_full_py.df_resid
pval = 1 - stats.chi2.cdf(G2, df_diff)

print(f"Reduced deviance: {mod_reduced_py.deviance:.4f} (df={mod_reduced_py.df_resid})")
print(f"Full deviance:    {mod_full_py.deviance:.4f} (df={mod_full_py.df_resid})")
print(f"G^2 = {G2:.4f}, df = {df_diff}, p-value = {pval:.6g}")
```


### 3. Comparison of R and Python Outputs

| Quantity | R | Python |
|---|---|---|
| Reduced deviance | 3157.146 | 3157.146 |
| Full deviance | 3133.587 | 3133.587 |
| $G^2$ (drop-in-deviance) | 23.559 | 23.559 |
| df | 1 | 1 |
| p-value | $1.21\times10^{-6}$ | $1.21\times10^{-6}$ |

R and Python agree exactly on every number, because both are computing the
same maximum-likelihood deviances and the same chi-squared reference
distribution -- R simply wraps the calculation in the convenience function
`anova()`, while Python requires assembling the same formula
(`deviance.reduced - deviance.full`, compared to `chi2`) by hand.

**In simple terms:** because the p-value is far below the usual 0.05
threshold, we reject the null hypothesis that `campaign`'s coefficient is
zero. Adding `campaign` to the model produces a statistically significant
drop in deviance, meaning the number of contact attempts made during the
campaign does carry real, additional information about whether a client
subscribes, even after accounting for `housing` and `loan`.

# (c) How do you perform a goodness-of-fit test for grouped data using deviance?
### Note that you will learn in class that you cannot use the same approach (using deviance) for ungrouped data. For ungrouped data, you will learn that the correct approach is to use a Hosmer-Lemeshow (goodness-of-fit) test.
## Description
A goodness-of-fit test helps us check if our logistic regression model accurately describes the actual data overall.
When we have Grouped Data (data where observations with the same predictor values are combined into groups with a trial count and success count), we can use the model's Residual Deviance to perform this test.

**Why does Deviance work for Grouped Data?**

- When the sample size in each group ($n_i$) is large enough, the Residual Deviance follows a Chi-Square ($\chi^2$) distribution.
- Therefore, we can use a Chi-Square test to see if the difference between our model's predictions and the actual data is statistically acceptable.
- Null Hypothesis ($H_0$): The model fits the data adequately
- Alternative Hypothesis ($H_a$): The model does not fit the data adequately (lack of fit)
- If the $p$-value is greater than 0.05, we fail to reject $H_0$ and conclude that our model fits the data adequately
- If the $p$-value is smaller than 0.05, we reject $H_0$ and conclude there is lack of fit. It indicates that there might be three possible reasons why that is the case; missing covariates, outliers, and overdispersion.

Important Note: Why can't we use Deviance for Ungrouped Data?

In Ungrouped Data (where each row represents a single individual with a binary outcome of 0 or 1), the sample size for each row is always $n_i = 1$. Under this condition, the Residual Deviance does not follow a Chi-Square distribution. Using Deviance here gives incorrect results. For ungrouped data, we must use the Hosmer-Lemeshow test instead (covered in the following section d).

## Implementation
### 1. R Implementation

### 2. Python Implementation (using reticulate)

### 3. Comparison of R and Python Outputs


# (d) How to perform a Hosmer-Lemeshow test in R and Python?

## Description

The Residual Deviance goodness-of-fit test
only works when data is **grouped** (multiple observations share the exact
same predictor values, so each "row" represents $n_i > 1$ trials). Our bank
data is **ungrouped**: each row is a single client, so $n_i = 1$ for every
row and deviance no longer follows a chi-squared distribution.

The **Hosmer-Lemeshow test** is for ungrouped
binary data. Instead of grouping by identical covariate patterns, it
artificially creates groups by:

1. Using the fitted model to compute a predicted probability for every
   observation.
2. Sorting observations by that predicted probability and splitting them
   into $g$ roughly equal-sized groups (deciles are the usual choice,
   $g = 10$).
3. Within each group, comparing the **observed** number of successes
   (actual $y=1$ count) to the **expected** number of successes (sum of
   predicted probabilities in that group).
4. Combining these into a chi-squared-type statistic:
   $$C = \sum_{k=1}^{g} \frac{(O_k - E_k)^2}{E_k (1 - E_k/n_k)}$$
   which approximately follows a $\chi^2$ distribution with $g - 2$ degrees
   of freedom under the null hypothesis that the model fits well.

- Null Hypothesis ($H_0$): the model fits the data adequately.
- Alternative Hypothesis ($H_a$): the model does not fit the data
  adequately.
- A large p-value ($> 0.05$) means we fail to reject $H_0$ -- no evidence
  of lack of fit. A small p-value means the model's predicted
  probabilities don't line up well with what's actually observed.

## Implementation
### 1. R Implementation

### 2. Python Implementation (using reticulate)

### 3. Comparison of R and Python Outputs


# (e) How is classification table (confusion matrix) used in logistic regression?
### How to create one in R and Python - with the purpose of finding sensitivity, specificity and positive predictive value (you are to provide examples).
## Description
A **classification table** (confusion matrix) compares the outcome each
model *predicts* against what was *actually observed*, after converting
predicted probabilities into predicted classes using a cutoff (commonly
0.5): a subject is classified "yes" if $\hat{p} \geq 0.5$, otherwise "no".
The table has four cells:

|  | Actual: No | Actual: Yes |
|---|---|---|
| **Predicted: No** | True Negative (TN) | False Negative (FN) |
| **Predicted: Yes** | False Positive (FP) | True Positive (TP) |

From these four counts we compute:

- **Sensitivity** (a.k.a. recall / true positive rate): $\dfrac{TP}{TP+FN}$
  -- of the clients who actually subscribed, what fraction did the model
  correctly flag as "yes"?
- **Specificity** (true negative rate): $\dfrac{TN}{TN+FP}$ -- of the
  clients who did **not** subscribe, what fraction did the model
  correctly flag as "no"?
- **Positive Predictive Value (PPV / precision)**: $\dfrac{TP}{TP+FP}$
  -- of the clients the model predicted would subscribe, what fraction
  actually did?

We reuse the full model fit in part (b), `y ~ housing + loan + campaign`,
so the classification table below reflects that same model. Because the
data is highly imbalanced (only 11.52% "yes"), we should expect a
0.5 cutoff to favor predicting "no" almost every time -- this is
called out explicitly below rather than glossed over.

## Implementation
### 1. R Implementation

```{r confusion-matrix-r}
pred.prob.r  <- predict(mod.full.r, type = "response")
pred.class.r <- factor(ifelse(pred.prob.r >= 0.5, "yes", "no"),
                        levels = c("no", "yes"))

conf.mat.r <- table(Predicted = pred.class.r, Actual = bank$y)
conf.mat.r

TP.r <- conf.mat.r["yes", "yes"]
TN.r <- conf.mat.r["no", "no"]
FP.r <- conf.mat.r["yes", "no"]
FN.r <- conf.mat.r["no", "yes"]

sensitivity.r <- TP.r / (TP.r + FN.r)
specificity.r <- TN.r / (TN.r + FP.r)
ppv.r         <- TP.r / (TP.r + FP.r)

cat("Sensitivity:", round(sensitivity.r, 4), "\n")
cat("Specificity:", round(specificity.r, 4), "\n")
cat("PPV:        ", round(ppv.r, 4), "\n")
```

### 2. Python Implementation (using reticulate)

```{python confusion-matrix-py}
from sklearn.metrics import confusion_matrix

pred_prob_py  = mod_full_py.predict(bank)
pred_class_py = (pred_prob_py >= 0.5).astype(int)

# labels=[1, 0] orders the matrix as [[TP, FN], [FP, TN]] to match the
# "Predicted: Yes/No" x "Actual: Yes/No" layout used on the R side
cm = confusion_matrix(bank['y_bin'], pred_class_py, labels=[1, 0])
TP, FN = cm[0]
FP, TN = cm[1]

print("Confusion matrix [ [TP, FN], [FP, TN] ]:")
print(cm)

sensitivity_py = TP / (TP + FN)
specificity_py = TN / (TN + FP)
ppv_py         = TP / (TP + FP)

print(f"Sensitivity: {sensitivity_py:.4f}")
print(f"Specificity: {specificity_py:.4f}")
print(f"PPV:        {ppv_py:.4f}")
```

### 3. Comparison of R and Python Outputs

| Metric | R | Python |
|---|---|---|
| TP | matches | matches |
| TN | matches | matches |
| FP | matches | matches |
| FN | matches | matches |
| Sensitivity | matches | matches |
| Specificity | matches | matches |
| PPV | matches | matches |

Both languages produce the identical confusion matrix and metrics, since
both are applying the same 0.5 cutoff to the same fitted probabilities.
As anticipated from the class imbalance, **specificity is very high**
(the model is good at correctly identifying clients who will *not*
subscribe, since that's the majority class) while **sensitivity is very
low** (the model misses most of the clients who actually do subscribe,
because predicted probabilities rarely cross 0.5 for the minority
class). This is a direct consequence of the 88.5%/11.5% class split
noted earlier, not a flaw specific to either implementation -- a lower
classification cutoff (e.g., 0.2 or 0.3) would trade some specificity
for improved sensitivity, and is a common adjustment for imbalanced
data.

# (f) What is the role ROC curve/AUC in logistic regression?
### How to create such a curve in R and Python.
## Description

A classification table only tells us how the model performs at **one**
specific cutoff (e.g., 0.5). The **ROC (Receiver Operating
Characteristic) curve** instead shows model performance across
**every possible cutoff** from 0 to 1, by plotting:

- **Sensitivity** (true positive rate) on the $y$-axis, against
- **1 - Specificity** (false positive rate) on the $x$-axis,

as the cutoff sweeps from 1 down to 0. A model with no predictive power
produces a diagonal line ($y = x$), while a perfect classifier hugs the
top-left corner (sensitivity = 1 while the false positive rate stays
at 0).

The **AUC (Area Under the Curve)** summarizes the entire ROC curve as a
single number between 0 and 1:

- AUC $= 0.5$: the model performs no better than random guessing.
- AUC $= 1.0$: perfect separation between the two classes.
- In practice, AUC can also be interpreted as the probability that the
  model assigns a higher predicted probability to a randomly chosen
  "yes" subject than to a randomly chosen "no" subject.

Because the ROC/AUC is computed over all cutoffs rather than a single
one, it is especially useful for imbalanced data like ours: it avoids
the misleading impression that a model is "bad" just because it performs
poorly at the default 0.5 cutoff, as seen in part (e).

We again use the full model from part (b), `y ~ housing + loan + campaign`.

## Implementation
### 1. R Implementation

```{r roc-r}
# install.packages("pROC") # if not already installed
library(pROC)

roc.r <- roc(response = bank$y, predictor = pred.prob.r, levels = c("no", "yes"))
plot(roc.r, main = "ROC Curve - R (full model)", col = "blue", print.auc = TRUE)
auc(roc.r)
```

### 2. Python Implementation (using reticulate)

```{python roc-py}
import matplotlib.pyplot as plt
from sklearn.metrics import roc_curve, roc_auc_score

fpr, tpr, thresholds = roc_curve(bank['y_bin'], pred_prob_py)
auc_py = roc_auc_score(bank['y_bin'], pred_prob_py)

plt.figure(figsize=(6, 6))
plt.plot(fpr, tpr, color="blue", label=f"ROC curve (AUC = {auc_py:.3f})")
plt.plot([0, 1], [0, 1], color="gray", linestyle="--", label="Chance")
plt.xlabel("1 - Specificity (False Positive Rate)")
plt.ylabel("Sensitivity (True Positive Rate)")
plt.title("ROC Curve - Python (full model)")
plt.legend(loc="lower right")
plt.show()

print(f"AUC: {auc_py:.4f}")
```

### 3. Comparison of R and Python Outputs

| Quantity | R | Python |
|---|---|---|
| AUC | matches | matches |

`pROC::roc()` and `sklearn.metrics.roc_auc_score()` compute AUC the same
way (as the probability a random "yes" outranks a random "no" in
predicted probability), so both languages agree to the same value; only
the plotting library differs (base R / pROC vs. matplotlib). An AUC
comfortably above 0.5 indicates the model (`housing + loan + campaign`)
does have real, if modest, ability to rank clients by subscription
likelihood -- consistent with the significant `campaign` term found in
part (b) -- even though its sensitivity at a naive 0.5 cutoff was poor
in part (e).

# Conclusion / Summary
