---
title: "Cross-Validation Techniques: A Comparative Study"
subtitle: "Using the Cleveland Heart Disease Dataset"
author: "Team Names Here"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(caret)
library(boot)
library(ggplot2)
library(knitr)
set.seed(2026)
```

# Introduction

# The Dataset

## Data Cleaning Pipeline

### Step 1: Load the raw data and add column headers

```{r step1-headers}
# The file downloaded from UCI has NO header row, so we set header = FALSE
# and assign the official column names ourselves. We also tell R to treat
# both "?" and blank cells as missing values (na.strings).
raw <- read.csv(r"(C:\Users\diegu\Downloads\cleveland.csv)", header = FALSE,
                 na.strings = c("?", ""))

colnames(raw) <- c("age", "sex", "cp", "trestbps", "chol", "fbs", "restecg",
                    "thalach", "exang", "oldpeak", "slope", "ca", "thal", "num")

cat("Raw data dimensions:", nrow(raw), "rows x", ncol(raw), "columns\n")
head(raw)
```

### Step 2: Handle missing values

```{r step2-missing}
# Only 'ca' and 'thal' contain missing values in this dataset
cat("Missing values per column:\n")
print(colSums(is.na(raw)))

# Only 6 of 303 rows are affected (~2%), so we drop those rows rather
# than impute, to keep things simple
raw <- na.omit(raw)
cat("\nRows remaining after removing missing values:", nrow(raw), "\n")
```

### Step 3: Prepare the target variable

```{r step3-target}
# 'num' is coded 0-4 (0 = no disease, 1-4 = increasing severity).
# We binarize it: 0 = no disease, 1 = disease present
raw$target <- ifelse(raw$num > 0, 1, 0)
raw$num <- NULL

# Turn target and sex into labeled factors for readability
raw$sex    <- factor(raw$sex,    levels = c(0, 1), labels = c("female", "male"))
raw$target <- factor(raw$target, levels = c(0, 1), labels = c("no_disease", "disease"))

heart <- raw

```

## Exploratory Data Analysis

# Technique 1: Data Split (Train/Test)

## The idea

## Implementation

## Comment

# Technique 2: Leave-One-Out Cross-Validation (LOOCV)

## The idea

## Implementation

## Comment

# Technique 3: k-fold Cross-Validation

## The idea

## Implementation

## Comment

# Technique 4: Repeated k-fold Cross-Validation

## The idea

## Implementation

## Comment


